---
title: "Biodiversity Modelling Summer School"
author:
- name: Dominique Gravel, Ph.D.
  affiliation: Université de Sherbrooke
- name: Andrew MacDonald, Ph.D.
  affiliation: Université de Sherbrooke
date: "2021-08-16"
description: |
  Evaluating ecological models with data
output:
  distill::distill_article:
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Learn more about creating websites with Distill at:
# https://rstudio.github.io/distill/website.html

# Learn more about publishing to GitHub Pages at:
# https://rstudio.github.io/distill/publish_website.html#github-pages
```

## Probabilities and distributions  

**16 August**  

know your data, define your problem.

```{r share-again, echo=FALSE}
xaringanExtra::use_share_again()
```

```{r slides-intro, echo=FALSE, fig.cap="[Slides](slides/index.html)"}
xaringanExtra::embed_xaringan(url = "slides/01_introduction.html", ratio = "16:9")
```


```{r slides-definitions, echo=FALSE, fig.cap="[Slides](slides/index.html)"}
xaringanExtra::embed_xaringan(url = "slides/02_definitions.html", ratio = "16:9")
```


```{r slides-rules, echo=FALSE, fig.cap="[Slides](slides/index.html)"}
xaringanExtra::embed_xaringan(url = "slides/03_rules.html", ratio = "16:9")
```


```{r slides-problem, echo=FALSE, fig.cap="[Slides](slides/index.html)"}
xaringanExtra::embed_xaringan(url = "slides/04_problem.html", ratio = "16:9")
```


```{r slides-solutions, echo=FALSE, fig.cap="[Slides](slides/index.html)"}
xaringanExtra::embed_xaringan(url = "slides/05_solutions.html", ratio = "16:9")
```

### Univariate probability distributions


```{r slides-distributions, echo=FALSE, fig.cap="[Slides](slides/index.html)"}
xaringanExtra::embed_xaringan(url = "slides/06_distributions.html", ratio = "16:9")
```


#### Continuous distributions

| distribution | positive or negative?       | Quick description                                              |
|--------------|-----------------------------|----------------------------------------------------------------|
| Normal       | positive or negative values | Can result from many small effects added together              |
| Lornormal    | positive values only        | results from many things multiplied together                   |
| Gamma        | positive values only        | can be the time to wait for a given number of things to happen |
| Exponential  | Positive only               | Lengths of time between random events                          |
| Beta         | Positive only               | Between 0 and 1. Can be any proportion.                        |

#### Discrete distributions


| distribution      | positive or negative? | Quick description                                                                                                     |
|-------------------|-----------------------|-----------------------------------------------------------------------------------------------------------------------|
| Poisson           | Positive              | Counting things that occur randomly over time                                                                         |
| Negative Binomial | Positive              | Counts of things as in the Poisson, but more variable. Also: the number of trials till a certain number of successes. |
| Binomial          | Positive              | Number of "successes" out of a number of trials, when probability of success is always the same                       |



## Maximum likelihood methods  

**18 August**  

write down equations for your model.


```{r slides-likelihood, echo=FALSE, fig.cap="[Slides](slides/index.html)"}
xaringanExtra::embed_xaringan(url = "slides/07_likelihood.html", ratio = "16:9")
```


```{r slides-likelihood-problem, echo=FALSE, fig.cap="[Slides](slides/index.html)"}
xaringanExtra::embed_xaringan(url = "slides/08_likelihood_problem.html", ratio = "16:9")
```


```{r slides-likelihood-solution, echo=FALSE, fig.cap="[Slides](slides/index.html)"}
xaringanExtra::embed_xaringan(url = "slides/09_likelihood_solution.html", ratio = "16:9")
```

## Optimization

```
August 20	Optimization algorithms: reveal the capacities of computers to solve complex problems numerically.
August 23	Bayesian statistics: how to properly represent uncertainty in ecological models.
August 25	Sampling algorithms: evaluating posterior distributions by your own.
```
